\documentclass{article}
\input{preamble-light.tex}
\title{Basic Fluid Dynamics}
\author{T. J. Crawford, J. Goedecke, P. Haas, E. Lauga, J. R. Lister, J. Munro, J. M. F. Tsang}

\begin{document}
\maketitle

\section{Relevant courses}

The relevant Cambridge undergraduate courses are IB Fluid Dynamics, II Fluid
Dynamics and II Waves. 

\section{Books}

\section{Notes}

\subsection{Lagrange, Euler and the material derivative}

There are two natural ways to describe a fluid flow. In the \textit{Eulerian
view}, one considers the evolution of fluid properties (such as velocity and
pressure) at a fixed point, while in the \textit{Lagrangian view}, one follows
the path of a certain fluid element and considers what happens to that element.

If the velocity field is $\bs{u}(\bs{x},t)$, the \textit{pathline} of a fluid
particle is the path traced out by that particle; it is the solution to 
$$ \dod{\bs{x}}{t} = \bs{u}(\bs{x},t). $$
This is useful in the Lagrangian description.

However, it is usually easier to work under the Eulerian view, and to take a
`snapshot' of fluid properties at all places at a given instant in time. A
\textit{streamline} is a curve that is instantaneously everywhere parallel to
the velocity field. 

For a time-independent fluid flow, streamlines and pathlines coincide with each
other, but this is not true for a time-dependent flow.

Consider a quantity $F(\bs{x},t)$.  What is the rate of change
of $F$ following the fluid particle at $(\bs{x},t)$? Since the local velocity is
$\bs{u}(\bs{x},t)$, the particle will move by a distance
$\delta\bs{x}=\bs{u}(x,t) \delta t$ over the time $\delta t$. Hence the change
in $F$ is
$$ \delta F = F(\bs{x}+\delta\bs{x},t+\delta t) - F(\bs{x},t)
     = (\delta\bs{x}\cdot\grad)F + \delta t \dpd{F}{t} + \cdots $$
and we define 
$$ \DDt{F}{t} = \lim_{\delta t\rightarrow0} 
 \frac{F(\bs{x}+\delta\bs{x},t+\delta t) - F(\bs{x},t)}{\delta t} 
 = \dpd{F}{t} + \bs{u}\cdot\grad F. $$
We call $\DDt{F}{t}$ the \textit{material derivative} or \textit{Lagrangian
derivative} of $F$, and we contrast it against $\pd{F}{t}$, which refers to the rate
of change of $F$ at a fixed position.

\subsection{Governing equations}

Mass conservation: 
$$ \dpd{\rho}{t} + \divg(\rho\bs{u}) = 0 $$

Momentum balance:
$$ \dpd{}{t}(\rho\bs{u}) + \divg(\rho\bs{u}\bs{u}) = \divg\bs{\sigma} + \bs{f} $$
or, using the mass conservation equation,
$$ \rho\left(\dpd{\bs{u}}{t} + \bs{u}\cdot\grad\bs{u}\right) = \divg\bs{\sigma} + \bs{f} $$
Here, $\bs{f}$ represents a \textit{body force}, such as gravity or a Coriolis
force.  And $\sigma$ is the \textit{Cauchy stress tensor} and describes the
effects of \textit{surface forces}. If $\dif S$ is a small surface element with
normal $\bs{n}$, then the force on that surface element by the fluid is
$\dif\bs{F}=\bs{\sigma}\cdot\bs{n}\dif S$, and the \textit{stress} (force per
unit area) is $\bs{\sigma}\cdot\bs{n}$.

Energy, assuming incompressibility:
$$ \dpd{}{t} \left(\frac{1}{2}\rho|\bs{u}|^2\right) 
   + \bs{u}\cdot\grad\left(\frac{1}{2}\rho|\bs{u}|^2\right) 
   = \divg(\bs{u}\cdot\bs{\sigma}) + \bs{u}\cdot\bs{f} - \bs{\sigma}:\bs{e} $$
where $\bs{e}$ is the \textit{rate of strain tensor}. The last term,
$\bs{\sigma}:\bs{e}$, represents \textit{dissipation} and is usually written as
$\Delta$.

One gets equations for mass and momentum conservation by applying the divergence
theorem on a small volume. A similar argument on angular momentum shows that,
assuming that long-range forces exert no couple on individual fluid elements,
the stress tensor must be symmetric: $\sigma_{ij}=\sigma_{ji}$.

The above holds for all incompressible fluids. To close the system, we need to
provide a relationship between $\bs{\sigma}$ and the other quantities. Such a
relationship is a \textit{constitutive equation}, and is determined by the
nature of the material.

An incompressible Newtonian fluid satisfies $\DDt{\rho}=\divg\bs{u}=0$ and has
the constitutive equation
$$ \bs{\sigma} = -p \bs{I} + 2\mu\bs{e}, $$
where $p$ is the \textit{pressure}. We do not need to provide another equation
to specify $p$ and close the system; $p$ is determined by the condition that
$\divg\bs{u}=0$.\footnote{In some sense, $p$ acts as a Lagrange multiplier which
imposes the incompressibility constraint.} The strain rate tensor is given by
$$ \bs{e} = \frac{1}{2}(\grad\bs{u}+\grad\bs{u}^T) $$
so that $\trace\bs{e} = 0$. And the dissipation rate is given by
$$\Delta = \bs{\sigma}:\bs{e}=2\mu\bs{e}:\bs{e}.$$

The assumptions that give rise to this constitutive relation are that the the
deviatoric stress (i.e. the anisotropic part of $\bs{\sigma}$) is
\textit{linear} in $\grad\bs{u}$, and \textit{instantaneous} (i.e. there is no
dependence on the history of deformation). 

After all this, one gets the \textit{Navier-Stokes equations}:
$$ \rho\DDt{\bs{u}} = -\grad p + \mu\grad^2 \bs{u} $$
and 
$$ \divg\bs{u} = 0. $$

\subsection{Boundary conditions}

\paragraph{Kinematic boundary conditions}  At a fluid-fluid interface,
$\bs{u}\cdot\bs{n}$ is continuous (by mass conservation) and
$\bs{u}\cross\bs{n}$ is continuous (to avoid infinite stresses). 

At a rigid boundary moving with velocity $\bs{U}$, the kinematic BCs simplify to
$$ u = U. $$

For a surface given by $z=h(x,y,t)$, the kinematic BCs can be expressed as
$$ \DDt{}(h-z) = 0 $$
on $z=h$.

\paragraph{Dynamic boundary conditions} In the absence of surface tension,
$\bs{\sigma}\cdot\bs{n}$ is continuous. With surface tension,
$$ [\bs{\sigma}\cdot\bs{n}] = \gamma\kappa\bs{n} - \grad_s\gamma $$
where $\gamma$ is the \textit{coefficient of surface tension}, $\grad_s$ is the
gradient on the surface, and $\kappa = \grad_s\cdot\bs{n}$ is the
\textit{total curvature} of the surface.

\section{Exercises}

\end{document}
